---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

#### 1. Loading the data
```{r read_data}
data <- read.csv("activity.csv", header = TRUE)
```

#### 2. Checking the data structure
```{r data_structure, results="hide", warning = FALSE}
- head(data)
- tail(data)
- dim(data)
```

#### 3. Formatting the dates
```{r format_dates}
data$date <- as.Date(data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

#### 1. Calculating the total number of steps taken each day
```{r tot_steps_day}
stepsbyday <- aggregate(steps ~ date, data = data, sum)
```

#### 2. Making the histogram of the data
```{r hist_data}
hist(stepsbyday$steps, col = 8, main = "Histogram of Number of Steps Taken per Day", 
     xlab = "Total Number of Steps")
```

#### 3. Calculating Mean and Median of total number of steps taken per day
```{r mean_median_tot_steps_day}
meanSteps <- format(mean(stepsbyday$steps, na.rm = TRUE))

medianSteps <- format(median(stepsbyday$steps, na.rm = TRUE))
```
The mean total number of steps taken per day is `r meanSteps` and the median is `r medianSteps`.

## What is the average daily activity pattern?

#### 1. Calculating the average steps taken
```{r avg_steps}
meanInterval <- aggregate(steps~interval, data, mean, na.rm = TRUE)
```

#### 2. Creating the time series plot
```{r time_series_plot}
plot(meanInterval$interval, meanInterval$steps, type = "l", 
     main = "Average Number of Steps Taken by 5-Minute Interval", 
     xlab = "5-Minute Interval",
     ylab = "Average Number of Steps Across all Days")
```

#### 3. Calculating the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps
```{r 5_min_interval}
maxSteps <- round(max(meanInterval$steps),1)

maxInterval <- meanInterval$interval[which.max(meanInterval$steps)]
```
The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is 835.

## Imputing missing values

#### 1. Total number of missing values in the dataset
```{r tot_missing_values}
missingRecords <- sum(is.na(data$steps))
```

#### 2. Replace missing step values with the mean for the entire observation period using the average for the entire observation period
```{r replace_missing_values}
stepfill = vector()
for (item in data$steps) {
        if (is.na(item)) { # if a step is missing
                stepfill = append(stepfill, mean(data$steps, na.rm = TRUE)) # return the mean 
        } else {
                stepfill = append(stepfill, item)
        }        
}
```
The strategy used to replicate missing data was to calculate average number of steps for the entire observation period.


#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r dataset_missing_filled}
dataFull <- data
dataFull$steps <- stepfill
```

#### 4. Calculating the total number of steps taken each day
```{r tot_steps_each_day}
stepsbydayNew = aggregate(steps ~ date, data = dataFull, sum)
```

#### 5. Histogram of the total number of steps taken each day
```{r hist_tot_steps_day}
hist(stepsbydayNew$steps, col = 8, main = "Number of Steps Taken per Day (Adjusted)",
     xlab = "Total Number of Steps")
```

#### 6. Calculating the mean and median total number of steps taken per day
```{r new_mean_median}
newMean <- format(mean(stepsbydayNew$steps, na.rm = TRUE)) 
newMedian <- format(median(stepsbydayNew$steps, na.rm = TRUE))
```

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Setting the date format
```{r set_dates}
data$date <- as.Date(data$date, "%Y-%m-%d")
```

#### 2. Adding a new column indicating day of the week 
```{r day_of_week_col}
data$day <- weekdays(data$date)
```

#### 3. Adding a new column for day_type to identify the weekday
```{r set_weekday}
data$day_type <- c("weekday")
```

#### 4. Making the day_type as weekend if the day is Saturday or Sunday
```{r set_weekend}
for (i in 1:nrow(data)) {
        if (data$day[i] == "Saturday" || data$day[i] == "Sunday"){
                data$day_type[i] <- "weekend"
        }
}
```

#### 5. Changing the day_type data type
```{r}
data$day_type <- as.factor(data$day_type)
```

#### 6. Adding the mean steps in the day_type intervals 
```{r day_type_interval}
interval_steps <- aggregate(steps ~ interval + day_type, data, mean)
```

#### 7. Loading ggplot2 for plotting the results
```{r load_ggpllot2}
library(ggplot2)
```

#### 8. Plotting the results
```{r plot_results}
qplot(interval, steps, data = interval_steps, geom = c("line"), xlab = "Five Minute Interval", 
      ylab = "Number of steps", main = "Average Number of Steps Taken by Day Type") + facet_wrap( ~ day_type, ncol=1)
```
